MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-14 docs/paper/out/paper.tex
add my fixes/enhancements to the Introduction -- paper

Genome researchers have raised the alarm over big data in the past
\citep{hayden2015genome}, but even a more serious challenge might be
faced with the metagenomics boom. If we compare metagenomics data with
other genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling of the same patient along several weeks (or
years) is being used for the follow up of some intestinal pathologies,
for studying the evolution of the gut microbiome after antibiotic
treatment, or for colon cancer early detection
\citep{zeller2014potential}. This need of sampling across time adds more
complexity to metagenomics data storage and demands adapted algorithms
to detect state variations across time as well as idiosyncratic
commonalities of the microbiome of each individual
\citep{franzosa2015identifying}. In addition to the intra-individual
sampling-time dependence, metagenomic clinical test results vary
depending on the specific region of extraction of the clinical specimen.
This local variability adds complexity to the analysis since different
localizations (different tissues, different anatomical regions, healthy
or tumour tissues) are required to have a sufficiently complete
landscape of the human microbiome. Moreover, reanalysis of old samples
using new tools and better reference databases might be also demanded
from time to time.

During the last years other sciences as astronomy or particle physics
are facing the big data challenge but, at least, these science have
standards for data processing \citep{stephens2015big}. Global standards
for converting raw sequence data into processed data are not yet well
defined in metagenomics and there are shortcomings derived from the fact
that many bioinformatics methodologies currently used for metagenomics
data analysis were designed for a scenario very different that the
current one. These are some of the aspects that have suffered crucial
changes and advances with a direct impact in metagenomics data analysis:

\begin{enumerate}
\def\labelenumi{\roman{enumi}.}
\tightlist
\item
  The first aspect is related to the sequences to be analyzed: the reads
  are larger, the sequencing depth and the number of samples of each
  project are considerably bigger. The first metagenomics studies were
  very local projects, while nowadays the most fruitful studies are done
  at a global level (international, continental, national). This kind of
  global studies has yielded the discovery of clinical biomarkers for
  diseases of the importance of cancer, obesity or inflammatory bowel
  diseases and has allowed exploring the biodiversity in many earth
  environments
\item
  The second aspect derives from the impressive genomics explosion, its
  effect being felt in this case in the reference sequences. The immense
  amount of sequences available in public repositories demands new
  approaches in curation, update and storage for metagenomics reference
  databases: current models will or already have problems to face the
  future avalanche of metagenomic sequences.
\item
  The third aspect to consider for metagenomics data analysis is related
  to the appearance of new models for massive computation and storage
  such as the so-called cloud, or the widespread adoption of programming
  methodologies like functional programming, or, more speculatively,
  dependently typed programming. The immense new possibilities that
  these advances offer must have a direct impact in metagenomics data
  analysis.
\item
  And finally the new social manner to do science, and especially
  genomic science is the fourth aspect to consider. Metagenomics evolves
  in a social and global scenario following a science democratization
  trend in which many small research groups from distant countries share
  a common big metagenomics project. This global cooperation demands
  systems allowing following exactly the same pipelines using equivalent
  cloud resources to modularly execute the analysis in an asynchronous
  way of working between different groups. This new scenario calls for
  new methods and tools to handle the current and future volume of
  metagenomic data with the sufficient speed of analysis.
\end{enumerate}

>>>>

Genome researchers have raised the alarm over big data in the past
\citep{hayden2015genome}, but even a more serious challenge might be
faced with the metagenomics boom. If we compare metagenomics data with
other genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling of the same patient along several weeks (or
years) is being used for the follow up of some intestinal pathologies,
for studying the evolution of the gut microbiome after antibiotic
treatment, or for colon cancer early detection
\citep{zeller2014potential}. This need of sampling across time adds more
complexity to metagenomics data storage and demands adapted algorithms
to detect state variations across time as well as idiosyncratic
commonalities of the microbiome of each individual
\citep{franzosa2015identifying}. In addition to the intra-individual
sampling-time dependence, metagenomic clinical test results vary
depending on the specific region of extraction of the clinical specimen.
This local variability adds complexity to the analysis since different
localizations (different tissues, different anatomical regions, healthy
or tumour tissues) are required to have a sufficiently complete
landscape of the human microbiome. Moreover, re--analysis of old samples
using new tools and better reference databases might be also demanded
from time to time.

Other disciplines such as astronomy or particle physics have faced the
big data challenge before. A key difference is the existence of
standards for data processing \citep{stephens2015big}; in metagenomics
global standards for converting raw sequence data into processed data
are not yet well defined, and there are shortcomings derived from the
fact that most bioinformatics methodologies used for metagenomics data
analysis were designed for scenarios very different from the current
one. These are some of the aspects that have suffered crucial changes
and advances with a direct impact in metagenomics data analysis:

\begin{enumerate}
\def\labelenumi{\roman{enumi}.}
\tightlist
\item
  \textbf{Sequence data} the reads are larger, the sequencing depth and
  the number of samples of each project are considerably bigger. The
  first metagenomics studies were very local projects, while nowadays
  the most fruitful studies are done at a global level (international,
  continental, national). This kind of global studies has yielded the
  discovery of clinical biomarkers for diseases of the importance of
  cancer, obesity or inflammatory bowel diseases and has allowed
  exploring the biodiversity of varied earth environments.
\item
  \textbf{The genomics explosion} its effect being felt in this case in
  the reference sequences. The immense amount of sequences available in
  public repositories demands new strategies for curation, update and
  storage of metagenomics reference databases: current models will
  (already) have problems to face the future avalanche of metagenomic
  sequence data.
\item
  \textbf{Cloud computing} the appearance of new models for massive
  computation and storage such as the so-called cloud, or the widespread
  adoption of programming methodologies like functional programming, or,
  more speculatively, dependently typed programming. The immense new
  possibilities that these advances offer must have a direct impact in
  metagenomics data analysis.
\item
  \textbf{Open science} the new social manner to do science,
  particularly so in the case of genomics, brings its own set of
  requirements. Metagenomics evolves in a social and global scenario
  following a science democratization trend in which many small research
  groups from distant countries share a common big metagenomics project;
  this global cooperation demands systems allowing for reproducible data
  analysis, data interoperability, and tools and practices for
  asynchronous collaboration between different groups.
\end{enumerate}
